Handling String-Number Conversion in String Constraint Solving

### 2019-11-20 draft.tex
update evaluation and minor update to intro
add experimental results to introduction

%As solvers are applied to symbolic execution of programming languages, this is increasingly becoming a severe limitation. For many languages such as Java, conversion between integers and strings is a niche operation, and much symbolic execution can be done without supporting it; however, this is not true of some scripting languages, of which JavaScript is the most notable. JavaScript is a key language---it powers most interactive content on the Web, and increasingly server-side code with Node.js---and it has conversions between strings and integers embedded in the core of its semantics.

\begin{verbatim}
for(var i = 0; i < 10; i++) {
 arr[i] = 0
}
\end{verbatim}

@@

\hspace{-4mm}\begin{tabular}{l|l|c}
1&	{\tt{x = [0, 0, 0, 0, 0]}} & [0, 0, 0, 0, 0] \\
2&	{\tt{x[3] = 4}} & [0, 0, 0, 4, 0] \\
3&	{\tt{x[03] = 2}} & [0, 0, 0, 2, 0] \\
4&	{\tt{x["3"] = 5}} & [0, 0, 0, 5, 0] \\
5&	{\tt{x["03"] = 7}} & [0, 0, 0, 5, 0] and x["03"] = 7\\
6&	{\tt{x["03"-1] = 2}} & [0, 0, 2, 5, 0] and x["03"] = 7\\
\end{tabular}

@@

In this paper, we propose a framework that efficiently handles string constraints with string-number conversion. Due to the diffifulty of solving the problem exactly, our framework combines over and under-approximation of string constraints. The over-approximation is for proving UNSAT when possible, while the under-approximation is for proving SAT when possible. Both over- and under-approximation fall in a decidable fragment of string constraints that can we can efficiently solve.

@@

Our decision procedure has two steps: the first step is over-approximating the set of input constraints into a set that falls in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Observe that we could  over-approximate the input constraint into any decidable fragment, e.g. the acyclic fragment~\cite{abdulla2014string} or the straight-line fragment~\cite{chen2019decision}. Our choice of the  chain-free fragment~\cite{abdulla2019chain} is motivated by the fact that the chain-free fragment is the {\em largest} known decidable fragment for that class of string constraints. In our example, we over-approximate the formula $\Phi$ by converting $xy=yx$ to two formulae $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1 \vee (n\neq -1 \wedge x\in [0-9]^*$). The over-approximation module uses also some heuristics to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some case, we could derive the value of $n=12$ from some integer constraints.  Then we could further derive the value of $x$ is in the regular language $(0^*12)$. 
Observe that if the over-approximation is UNSAT then our decision procedure  declares that the original formula is also UNSAT and terminates. Since $\Phi$ is satisfiable, the over-approximation module will not return UNSAT.

@@

The second step of our decision procedure is only enabled if the over-approximation is SAT. In this case, our decision procedure restricts the domain of each string variable to strings that obey some predefined and parameterized pattern.  Adjusting the parameters enlarges or prunes the potential solution space. In this paper, we propose to use patterns defined by \emph{parametric flat automata} (PFA). A PFA is a {\em flat} finite state automaton consisting of a predefined sequence of loops, each of fixed length (see Figure \ref{fig:sfa_def}). The approach based on PFA is very flexible yet allows very efficient manipulation. One can pick PFA of \emph{arbitrary structure} for search space restriction and reduce the string constraint solving problem to a linear formula satisfiability problem in polynomial-time. Furthermore, we associate a unique natural number with each character. This allows us to assume that the input alphabet is a subset of natural numbers to avoid the \textit{alphabet explosion problem} from which the approach in~\cite{abdulla2017flatten} suffers.  We label each transition between two states of a PFA with a unique \emph{character} variable (whose domain is the set of natural numbers) instead of having a transition between each two states for each symbol in the alphabet.  
Such reduction based on PFA is the key to handling string-number conversion efficiently. When we restrict the solution space to those defined by PFAs, the original undecidable problem (i.e., string constraint with string-number conversion) is reduced to the satisfiability problem of linear formulae and can be solved efficiently.

@@

If a model of the produced linear formula is found, then the procedure concludes SAT with an assignment to the string variables. If not, our procedure changes the PFA instages to be more expressive (by adding more states and transitions) to systematically search for the solution. Depending on the implementation, one may choose to report unknown after some certain steps.

@@

To demonstrate the usefulness of our approach, we have implemented our decision procedure in an open source solver, called {\sc } ....
\todo{Talk about experiments}

>>>>

%As solvers are applied to symbolic execution of programming languages, this is increasingly becoming a severe limitation. For many languages such as Java, conversion between integers and strings is a niche operation, and much symbolic execution can be done without supporting it; however, this is not true of some scripting languages, of which JavaScript is the most notable. JavaScript is a key language---it powers most interactive content on the Web, and increasingly server-side code with Node.js---and it has conversions between strings and integers embedded in the core of its semantics. 
\smallskip
\begin{center}
	\begin{minipage}{5cm}
		\begin{verbatim}
    for(var i = 0; i < 10; i++) {
        arr[i] = 0
    }
		\end{verbatim}
	\end{minipage}
\end{center}
\smallskip

@@

\medskip
\hspace{-4mm}\begin{tabular}{l|l|c}
1&	{\tt{x = [0, 0, 0, 0, 0]}} & [0, 0, 0, 0, 0] \\
2&	{\tt{x[3] = 4}} & [0, 0, 0, 4, 0] \\
3&	{\tt{x[03] = 2}} & [0, 0, 0, 2, 0] \\
4&	{\tt{x["3"] = 5}} & [0, 0, 0, 5, 0] \\
5&	{\tt{x["03"] = 7}} & [0, 0, 0, 5, 0] and x["03"] = 7\\
6&	{\tt{x["03"-1] = 2}} & [0, 0, 2, 5, 0] and x["03"] = 7\\
\end{tabular}
\medskip

@@

In this paper, we propose a framework that efficiently handles string constraints with string-number conversion. Due to the difficulty of solving the problem exactly, our framework combines over and under-approximation of string constraints. The over-approximation is for proving UNSAT when possible, while the under-approximation is for proving SAT when possible. Both over- and under-approximation fall in a decidable fragment of string constraints that can we can efficiently solve.

@@

Our decision procedure has two steps: the first step is over-approximating the set of input constraints into a set that falls in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Observe that we could  over-approximate the input constraint into any decidable fragment, e.g. the acyclic fragment~\cite{abdulla2014string} or the straight-line fragment~\cite{chen2019decision}. Our choice of the  chain-free fragment~\cite{abdulla2019chain} is motivated by the fact that the chain-free fragment is the {\em largest} known decidable fragment for that class of string constraints. In our example, we over-approximate the formula $\Phi$ by converting $xy=yx$ to two formulas $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1 \vee (n\neq -1 \wedge x\in [0-9]^*$). The over-approximation module uses also some heuristics to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some case, we could derive the value of $n=12$ from some integer constraints.  Then we could further derive the value of $x$ is in the regular language $(0^*12)$. 
Observe that if the over-approximation is UNSAT then our decision procedure  declares that the original formula is also UNSAT and terminates. Since $\Phi$ is satisfiable, the over-approximation module will not return UNSAT.

@@

The second step of our decision procedure is only enabled if the over-approximation is SAT. In this case, our decision procedure restricts the domain of each string variable to strings that obey some predefined and parameterized pattern.  Adjusting the parameters enlarges or prunes the potential solution space. In this paper, we propose to use patterns defined by \emph{parametric flat automata} (PFA). A PFA is a {\em flat} finite state automaton consisting of a predefined sequence of loops, each of fixed length (see Figure \ref{fig:sfa_def}). The approach based on PFA is very flexible yet allows very efficient manipulation. One can pick PFA of \emph{arbitrary structure} for search space restriction and reduce the string constraint solving problem to a linear formula satisfiability problem in polynomial-time. Furthermore, we associate a unique natural number with each character. This allows us to assume that the input alphabet is a subset of natural numbers to avoid the \textit{alphabet explosion problem} from which the approach in~\cite{abdulla2017flatten} suffers.  We label each transition between two states of a PFA with a unique \emph{character} variable (whose domain is the set of natural numbers) instead of having a transition between each two states for each symbol in the alphabet.  
Such reduction based on PFA is the key to handling string-number conversion efficiently. When we restrict the solution space to those defined by PFAs, the original undecidable problem (i.e., string constraint with string-number conversion) is reduced to the satisfiability problem of linear formulas and can be solved efficiently.

@@

If a model of the produced linear formula is found, then the procedure concludes SAT with an assignment to the string variables. If not, our procedure changes the PFAs to be more expressive ones (by adding more states and transitions) to systematically search for a solution. Depending on the implementation, one may choose to report unknown after some certain steps.

@@

To demonstrate the usefulness of our approach, we have implemented our decision procedure in an open source solver, called {\tool} and evaluated it on a large set of benchmarks obtained from literatures and from symbolic execution of real world programs. The experimental results show that {\tool} is among the best tools for solving basic string constraints and significantly outperforms all other tools on benchmarks with string-number conversion constraints. In this benchmark, the total amount of tests cannot be solved by {\tool} is only a half to the second best tool.

\todo{Talk about experiments}
